---
title: "Matching Darien 2026 species names to WCVP"
author: "Antoine Caron-Guay, Helene Muller-Landau, Paula Uzcategui"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Match WCVP names with species name

**Input:** species list from Darien (2026)

**Workflow:**

1. Match species names against WCVP (World Checklist of Vascular Plants)

2. Identify and review potential mismatches:

   - A mismatch occurs when WCVP returns a name different from the original lookup name
   - Most mismatches arise from minor spelling differences between the original names
     and WCVP entries. Others occur because some infraspecies are not in WCVP. Rarer
     mismatches occur when WCVP fuzzy matching points to the wrong name entirely.

3. Correct identified mismatches and save list with WCVP columns

**Output:**

- `_WCVP13` — original species list columns plus WCVP matched and accepted name columns
- `_WCVP13_CHANGES` — species where the original name differs from the WCVP accepted name
- `_WCVP13_GBIF` — above with GBIF taxon IDs added for both the matched and accepted names

```{r libraries, echo=TRUE, message=FALSE, warning=FALSE, results="hide"}

# installing wcvp
#install.packages("remotes")
#library(remotes)
#remotes::install_github("matildabrown/rWCVP")

rm(list=ls())

library(readr)
library("dplyr")
library("tidyr")
library("stringr")
suppressPackageStartupMessages({
  library(rWCVP)
  library(rWCVPdata)
})
library(TNRS)
library(DT)
library(stringi)
library(rgbif)


citation("rWCVPdata") # according to this, this is using version 13.

options(DT.options = list(scrollX = TRUE, pageLength = 15))
```

```{r globalvars}
# Load WCVP backbone
wcvp_names <- rWCVPdata::wcvp_names

# INPUT PATHS
original_list_path <- "projects/2026_darien/Especies_Darien_JV_2026_05_11.csv"

# Extract file name without extension
original_list_name <- tools::file_path_sans_ext(basename(original_list_path))

# OUTPUT PATHS
out_wcvp_list <- paste0("projects/2026_darien/", original_list_name, "_WCVP13", ".csv")
out_wcvp_changes_list <- paste0("projects/2026_darien/", original_list_name, "_WCVP13_CHANGES", ".csv")
out_wcvp_gbif_list <- paste0("projects/2026_darien/", original_list_name, "_WCVP13_GBIF", ".csv")
```


## Prepare datasets

```{r preparedata}

origlist <- read_csv2(original_list_path)

# modify this depending on the original dataset.
origlist <- origlist %>%
  rename(
    orig_family = "Name_matched_accepted_family",
    orig_name = "Name_submitted",
    orig_author = "Canonical_author",
  ) %>%
  mutate(
    orig_name = ifelse(is.na(orig_name) | str_squish(orig_name) == "", NA, str_squish(orig_name)),

    # extract genus and species from original name
    orig_genus = word(orig_name, 1),
    orig_species = word(orig_name, 2),
    orig_infraspecies_rank = word(orig_name, 3),
    orig_infraspecies = word(orig_name, 4),

    # create binomial name column. Keeping the prefix "X" and "sp. aff."
    orig_binomial = str_squish(
      paste0(orig_genus, " ", orig_species)
  )) %>%
  filter(!is.na(orig_name)) %>%
  select(orig_family, orig_name, orig_author, orig_binomial, orig_genus, orig_species, orig_infraspecies_rank, orig_infraspecies, sp6) %>%
  arrange(orig_name)


# find duplicates
dups <- origlist %>% group_by(orig_name) %>% filter(n() > 1) %>% select(orig_name)

if(nrow(dups) > 0){
  knitr::kable(dups, caption="Duplicated names in original")
}

knitr::kable(origlist, caption="Complete original list")
```

Working with `r nrow(origlist)` species.

### Match WCVP names

```{r mergewcvp}
# Helper function to merge and format accepted names from WCVP backbone
matchaccepted_names <- function(matcheswcvp, wcvp_backbone){
  matcheswcvp %>%
    left_join(
      wcvp_backbone %>%
        select(
          plant_name_id,
          # rename immediately
          wcvp_accepted_ipni_id  = ipni_id,
          wcvp_accepted_powo_id  = powo_id,
          wcvp_accepted_family   = family,
          wcvp_accepted_name_rank = taxon_rank,
          wcvp_accepted_name_status = taxon_status,
          wcvp_lifeform_description = lifeform_description,

          # rename after the mutate for readability
          genus, species, infraspecific_rank, infraspecies,
          parenthetical_author, primary_author
        ),
      by = c("wcvp_accepted_id" = "plant_name_id")
    ) %>%
    mutate(
      wcvp_accepted_name = case_when(
        is.na(species) | species == "" ~ genus,
        !is.na(infraspecies) & infraspecies != "" ~ paste(genus, species, infraspecific_rank, infraspecies),
        TRUE ~ paste(genus, species)
      ),

      wcvp_accepted_authority =
        paste0(
          ifelse(!is.na(parenthetical_author), paste0("(", parenthetical_author, ") "),""),
          coalesce(primary_author, "")
        )
    ) %>%
    rename(
      wcvp_accepted_author = primary_author,
      wcvp_accepted_name_genus = genus,
      wcvp_accepted_name_species = species,
      wcvp_accepted_name_infraspecific_rank = infraspecific_rank,
      wcvp_accepted_name_infraspecies = infraspecies
    )
}

# --- 1. Run Match ---
matchresult <- wcvp_match_names(subset(origlist), wcvp_names, name_col = "orig_name", author_col = "orig_author")


# --- 2. Sanity Checks & Duplicate Resolution ---
# Ensure all species are accounted for

# Identify names that mapped to multiple 'Accepted' records
manyaccepted <- matchresult %>%
  group_by(orig_name) %>%
  filter(sum(wcvp_status == "Accepted", na.rm = TRUE) > 1) %>%
  ungroup()

# If duplicates exist, force use of the original ID to avoid grouping explosion
if (nrow(manyaccepted) > 0) {
  matchresult <- matchresult %>%
    mutate(wcvp_accepted_id = ifelse(orig_name %in% manyaccepted$orig_name, wcvp_id, wcvp_accepted_id))
}

# --- 3. Clean Match Results (1 row per species) ---
matchresult_clean <- matchresult %>%
  mutate(
    status_priority = case_when(
      wcvp_status == "Accepted"     ~ 1,
      wcvp_status == "Synonym"      ~ 2,
      wcvp_status == "Illegitimate" ~ 3,
      wcvp_status == "Unplaced"     ~ 4,
      TRUE                          ~ 5
    )
  ) %>%
  group_by(orig_name) %>%
  slice_min(status_priority, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  matchaccepted_names(wcvp_names)
```


### Review mismatches

```{r spot_mismatch}

mismatches <- matchresult_clean %>%
  filter(orig_name != wcvp_name | is.na(wcvp_name) | is.na(wcvp_accepted_name)) %>%
  mutate(
    # parse WCVP name parts
    wcvp_genus = word(wcvp_name, 1),
    wcvp_species = word(wcvp_name, 2),
    wcvp_infraspecies_rank = word(wcvp_name, 3),
    wcvp_infraspecies = word(wcvp_name, 4),

    # flag component-level mismatches
    genus_mismatch = !is.na(orig_genus) &
                      (orig_genus != wcvp_genus | is.na(wcvp_genus)),
    species_mismatch = !is.na(orig_species) &
                        (orig_species != wcvp_species | is.na(wcvp_species)),
    infraspecies_rank_mismatch = !is.na(orig_infraspecies_rank) &
                                  (orig_infraspecies_rank != wcvp_infraspecies_rank |
                                   is.na(wcvp_infraspecies_rank)),
    infraspecies_mismatch = !is.na(orig_infraspecies) &
                             (orig_infraspecies != wcvp_infraspecies |
                              is.na(wcvp_infraspecies))
  )


printmismatches <- mismatches %>%
  select(
    orig_name, wcvp_name, wcvp_accepted_name, match_similarity,
    genus_mismatch, species_mismatch, infraspecies_rank_mismatch, infraspecies_mismatch
  ) %>%
  arrange(match_similarity)

datatable(printmismatches)


colsprint = c("orig_name", "wcvp_name","wcvp_accepted_name")
```

#### Case 1:

```{r printmisses}
# names that did not match exactly
mismatchcase1 <- mismatches %>% filter(orig_name != wcvp_name)

datatable( mismatchcase1 %>% select(all_of(colsprint)),
           caption="full name mismatched")
```

In these cases the original spelling differs slightly from the name matched in WCVP.

#### Case 2:

```{r case2}
# full name matched except infraspecies rank
mismatchcase2 <- mismatches %>% filter( infraspecies_rank_mismatch & !genus_mismatch & !species_mismatch & !infraspecies_mismatch)

datatable( mismatchcase2 %>% select(all_of(colsprint)),
           caption="full name matched except infraspecies rank")
```

The mismatch is a difference only in infraspecific rank. The WCVP accepted name is used for the full match.

#### Case 4:

```{r case4}
# genus, species, and rank matched but infraspecies did not
mismatchcase4 <- mismatches %>% filter( !genus_mismatch & !species_mismatch & infraspecies_mismatch & !infraspecies_rank_mismatch)

datatable( mismatchcase4 %>% select(all_of(colsprint)), caption="genus, species and infraspecific rank matched but infraspecies name did not ")
```

In these cases the subspecies did not match in WCVP. The binomial name is used instead (infraspecies dropped).

#### Other cases:

```{r case5}
# species not covered by the previous cases
missrest <- mismatches %>%
  filter(!orig_name %in% c(mismatchcase1$orig_name, mismatchcase2$orig_name, mismatchcase4$orig_name))

datatable( missrest %>% select(all_of(colsprint)), caption="Species not considered in previous cases")
```


#### Mismatch summary

The following WCVP columns will be retained in the output.

```{r wcvpcolumns}

# WCVP columns to retain in the final output
wcvp_columns <- c(
        # columns of matched name
        "wcvp_name", "wcvp_authors", "wcvp_rank", "wcvp_status", "wcvp_id", "wcvp_ipni_id", "match_similarity",

        # columns of accepted name
        "wcvp_accepted_id","wcvp_accepted_powo_id", "wcvp_accepted_ipni_id",
        "wcvp_accepted_name", "wcvp_accepted_name_genus",  "wcvp_accepted_name_species",
        "wcvp_accepted_name_rank", "wcvp_accepted_family",
        "wcvp_accepted_name_infraspecific_rank", "wcvp_accepted_name_infraspecies",
        "wcvp_accepted_author", "wcvp_accepted_authority",
        "wcvp_lifeform_description"
        )
```

```{r fixingmismatches}
datatable(mismatches %>% select(orig_name, wcvp_name, wcvp_accepted_name),
          caption="Species after setting the wcvp matched name to a closer match")
```


## Joining mismatch fixes to the main list

```{r matcheshandling}

# --- 4. Handle matches ---

# apply mismatch corrections back to the matched results
inc <- with(matchresult_clean, orig_name %in% mismatches$orig_name)
if (any(inc, na.rm = TRUE)) {
  corrections <- mismatches %>%
    select(all_of(c("orig_name", wcvp_columns)))

  matchresult_clean <- rows_update(matchresult_clean, corrections,
                                   by = "orig_name")
}


# --- 5. Final Join and Rename ---
# join WCVP columns onto the original species list
processlist <- origlist %>%
  left_join(matchresult_clean %>%
              select(all_of(c("orig_name", wcvp_columns))),
            by = "orig_name") %>%
  rename(
    # rename matched columns
    wcvp_matched_name = wcvp_name,
    wcvp_matched_authors = wcvp_authors,
    wcvp_matched_rank = wcvp_rank,
    wcvp_matched_status=wcvp_status,
    wcvp_matched_plant_name_id = wcvp_id,
    wcvp_matched_ipni_id = wcvp_ipni_id,

    wcvp_accepted_plant_name_id =  wcvp_accepted_id,


  ) %>%
  mutate(
  origmatches = orig_name == wcvp_matched_name
  )

# Review Results
knitr::kable(processlist %>% count(wcvp_matched_status))

knitr::kable(subset(processlist,
             is.na(wcvp_accepted_name) | !wcvp_matched_status %in% c("Accepted", "Synonym") ) %>%
        select(orig_family,orig_name,wcvp_matched_status,wcvp_accepted_name),
      caption = "Species not listed as Accepted or Synonym in WCVP")


# Species matched only to the genus level

onlygenus_match <- processlist %>%
  filter((is.na(word(orig_name,2)) | word(orig_name,2) == "" ) |
           (!is.na(wcvp_accepted_name) & is.na(word(wcvp_accepted_name,2)))) %>% select(orig_name, wcvp_matched_name, wcvp_accepted_name)

knitr::kable(onlygenus_match, caption="Species that matched only to genus ")
```


## Save output of all species

Showing columns saved in the final dataset:

```{r saveoutput}

# Save species list with WCVP columns
processlist <- processlist %>% mutate(across(everything(), ~replace_na(., "")))
write_csv(processlist, out_wcvp_list)

print(paste(colnames(processlist), collapse=", "))
```

Saved `r nrow(processlist)` species with their names verified with WCVP.


## Save output of all changes

Taxa where the original name differs from the WCVP accepted name are saved as a separate file.

```{r saveoutput_changes}

list_wcvp_changes <- processlist %>% filter(orig_name != wcvp_accepted_name)

# Save species with name changes
list_wcvp_changes <- list_wcvp_changes %>% mutate(across(everything(), ~replace_na(., "")))
write_csv(list_wcvp_changes, out_wcvp_changes_list)
```

Saved `r nrow(list_wcvp_changes)` species where the original name differs from the WCVP accepted name.


# Add GBIF ID

GBIF IDs are queried using the WCVP matched name, which is expected to produce an exact match.

```{r addgbifid}

process_list_path <- "projects/2026_darien/Especies_Darien_JV_2026_05_11_WCVP13.csv"

processlist <- read_csv(process_list_path)

processlist <- processlist %>%
  mutate(row_id = row_number())


datatable(processlist %>%
  filter(is.na(wcvp_matched_name)) %>%
  select(orig_name),
  caption="Current name does not have WCVP match")


# Build name input from WCVP matched name
names_for_gbif <- processlist %>%
  filter(!is.na(wcvp_matched_name)) %>%
  distinct(wcvp_matched_name, .keep_all = TRUE) %>%
  mutate(full_name_auth = paste(wcvp_matched_name, coalesce(wcvp_matched_authors, ""))) %>%
  rename(scientificName = full_name_auth)

# Query GBIF backbone
gbif_backbone_matched <- name_backbone_checklist(names_for_gbif, kingdom = "Plantae")

# Check match quality
gbif_match_summary <- gbif_backbone_matched %>%
  count(matchType)

print(gbif_match_summary)

# How many got a usable GBIF ID?
gbif_backbone_matched %>%
  summarise(
    total = n(),
    exact_match = sum(matchType == "EXACT", na.rm = TRUE),
    fuzzy_match = sum(matchType == "FUZZY", na.rm = TRUE),
    no_match    = sum(matchType == "NONE", na.rm = TRUE)
  )

gbif_backbone_matched <- gbif_backbone_matched %>%
  group_by(verbatim_name) %>%
  mutate(
    match_priority = case_when(
      matchType == "EXACT" ~ 1,
      matchType == "HIGHERRANK" ~ 2,
      matchType == "VARIANT" ~ 3,
      TRUE ~ 4
    )
  ) %>%
  arrange(match_priority, desc(confidence)) %>%
  slice(1) %>%
  ungroup()


# non-exact matches (Variant or higher rank)
datatable(gbif_backbone_matched %>%
            filter(!matchType %in% c("EXACT")) %>%
            select(verbatim_name, scientificName),
          caption="Species not matching exactly (verbatim refers to the fileds in the original list)")

datatable(gbif_backbone_matched %>%
            filter(!status %in% c("ACCEPTED")) %>%
            select(usageKey, verbatim_orig_name, scientificName, matchType, verbatim_wcvp_matched_status, status,  verbatim_wcvp_accepted_name, acceptedScientificName),
          caption="Species current name in the list is not accepted according to GBIF (verbatim refers to the fileds in the original list)")
```

The matched name may be a synonym — WCVP v13 indicates this as well. It is retained as the current name in the list, since the accepted name has not yet been reviewed by the STRI botanists in Panama.

The GBIF taxon key and scientific name are added as columns for the matched name.

```{r addgbifmatchednames}
processlist <- processlist %>%
  mutate(
    row_id = as.character(row_id)
  ) %>%
  left_join(
    gbif_backbone_matched %>%
      select(
        row_id = verbatim_row_id,
        wcvp_matched_name_gbif_id = usageKey,
        wcvp_matched_name_gbif_scientificName = scientificName,
        wcvp_matched_name_gbif_matchtype = matchType,
        wcvp_matched_name_gbif_status = status,
        wcvp_matched_name_gbif_acceptedScientificName = acceptedScientificName
      ),
    by = "row_id"
  )


# species without a GBIF match for the matched name
datatable(processlist %>% filter(is.na(wcvp_matched_name_gbif_id)) %>% select(orig_name, wcvp_matched_name))
```


The same lookup is performed for the **accepted name.**

```{r addgbifidaccepted}
datatable(processlist %>%
  filter(is.na(wcvp_accepted_name)) %>%
  select(orig_name),
  caption="Current name does not have WCVP accepted name")


# Build name input from WCVP accepted name
names_for_gbif <- processlist %>%
  filter(!is.na(wcvp_accepted_name)) %>%
  mutate(full_name_auth = paste(wcvp_accepted_name, coalesce(wcvp_accepted_authority, ""))) %>%
  rename(scientificName = full_name_auth)

# Query GBIF backbone
gbif_backbone_accepted <- name_backbone_checklist(names_for_gbif, kingdom = "Plantae")

# Check match quality
gbif_match_summary <- gbif_backbone_accepted %>%
  count(matchType)

print(gbif_match_summary)

# How many got a usable GBIF ID?
gbif_backbone_accepted %>%
  summarise(
    total = n(),
    exact_match = sum(matchType == "EXACT", na.rm = TRUE),
    fuzzy_match = sum(matchType == "FUZZY", na.rm = TRUE),
    no_match    = sum(matchType == "NONE", na.rm = TRUE)
  )


gbif_backbone_accepted <- gbif_backbone_accepted %>%
  group_by(verbatim_name) %>%
  mutate(
    match_priority = case_when(
      matchType == "EXACT" ~ 1,
      matchType == "HIGHERRANK" ~ 2,
      matchType == "VARIANT" ~ 3,
      TRUE ~ 4
    )
  ) %>%
  arrange(match_priority, desc(confidence)) %>%
  slice(1) %>%
  ungroup()

# non-exact matches (Variant or higher rank)
datatable(gbif_backbone_accepted %>%
            filter(!matchType %in% c("EXACT")) %>%
            select(verbatim_name, scientificName),
          caption="Species not matching exactly (verbatim refers to the fileds in the original list)")

datatable(gbif_backbone_accepted %>%
            filter(!status %in% c("ACCEPTED")) %>%
            select(verbatim_orig_name, verbatim_wcvp_matched_status, status, verbatim_wcvp_accepted_name=verbatim_name, acceptedScientificName),
          caption="Species WCVP accepted name in the list is not accepted according to GBIF (verbatim refers to the fileds in the original list)")
```

Note that the WCVP accepted name can itself be a synonym according to GBIF, since GBIF aggregates names from multiple sources. The GBIF accepted name is added as an additional column.

The GBIF taxon key and scientific name are added as columns for the accepted name.

```{r addgbifacceptednames}
processlist <- processlist %>%
  mutate(full_name_auth = paste(wcvp_accepted_name, coalesce(wcvp_accepted_authority, ""))) %>%
  left_join(
    gbif_backbone_accepted %>%
      select(
        full_name_auth = verbatim_name,
        wcvp_accepted_name_gbif_id = usageKey,
        wcvp_accepted_name_gbif_scientificName = scientificName,
        wcvp_accepted_name_gbif_matchtype = matchType,
        wcvp_accepted_name_gbif_status = status,
        wcvp_accepted_name_gbif_acceptedScientificName = acceptedScientificName
      ),
    by = "full_name_auth"
  ) %>%
  select(-full_name_auth)
```

Species without a GBIF match for either the matched or accepted name:

```{r missinggbifid}
datatable(processlist %>%
            filter(is.na(wcvp_matched_name_gbif_id) | is.na(wcvp_accepted_name_gbif_id)) %>%
            select(orig_name, wcvp_matched_name, wcvp_accepted_name),
          caption="Species withouth a GBIF id"
            )
```


## Save output of all species with GBIF IDs

Showing columns saved in the final dataset:

```{r saveoutput_gbif}

# Save species list with WCVP and GBIF columns
processlist <- processlist %>% mutate(across(everything(), ~replace_na(., "")))
write_csv(processlist, out_wcvp_gbif_list)

print(paste(colnames(processlist), collapse=", "))
```
